Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2024-01-16 transmission.tex
Updates from Overleaf

As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
I fit a similar model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.

@@

Having fit the prevalence model, I draw 10,000\todo{check number} prevalence trajectories in each stratum from the approximate posterior distribution of the model.
For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
Considering the relationship between incidence and prevalence to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, equation implies that, for any stratum $\vec{P} = \matr{S} \vec{Z}$ where $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum, $\matr{S}$ is a matrix as defined below, and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
The matrix $\matr{S}$ needs to be defined based on the survival function to preserve the relationship between incidence and prevalence.
It is lower-triangular of the following form:
\begin{align}
    (\matr{S})_{i,j} = \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases}
\end{align}
Since the matrix is lower-triangular, the solution for $\vec{P}$ can be found efficiently using forward substitution\todo{cite forward substitution}.
Forward substitution is numerically stable unless the diagonal elements are close to 0.
Here, the diagonal is $S(1) = 1$ and hence the matrix is well-conditioned.
\todo{I explained this more clearly in my first-year report, look that up!}

@@

The transmission part of the model follows \textcite{birrellRealtime}.
A novel approach (\cref{SEIR:sec:observation}) links the transmission to the data.
In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, it includes both the prevalence process and the observation process (as defined in \cref{E-inc-prev}).

@@

\subsection{Methods}

I use the same CIS data as in \cref{E-backcalc}, again considering each region independently.
The model will be stratified only by region and age group, so the data are aggregated to these levels.
Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.

@@

Previous studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

@@

The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}.
A variety of different approaches can be taken to link a transmission model to observations.
In the context of CIS, this requires representing the prevalence and observation processes (introduced in \cref{E-inc-prev:sec:prevalence-process,E-inc-prev:sec:observation-process} respectively)
The method to do so is known as the observation model.
The formulation of this model is heavily dependent on the data available.
This means that there is a wide variety of models used.
Therefore, this section  explains the model I propose for fitting to the CIS prevalence data.

@@

A much more efficient option is to approximate this convolution by adding compartments to the ODEs.
This approach also allows the infected, pre-positive compartment, which I denote $p_0$, to be represented.
$p_0$ increases at the same rate as individuals flow into $e_1$.
These individuals then transition into a set of compartment with their transition rates chosen to match the duration of positivity estimates in \cref{E-imperf-test}, $p_1, \dots, p_l$.
These compartments do not have a biological interpretation, unlike the other compartments in the system, but are purely a mathematical tool so that the time spent in these compartments matches the duration distribution.
This is the same idea behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).

@@

The modelled proportion of individuals that are PCR positive in the can then be linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
In this parameterisation (detailed in \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.

@@

Priors for the parameters are given in \cref{SEIR:table:priors}.
The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore I used assumed values or strongly informative priors on them.
$\sigma_\epsilon$ and $\rho$ are given priors that penalise model complexity.
The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
The prior on $\rho$ favours a null model of a binomial likelihood.
The remaining priors had weakly informative priors.

>>>>

As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
I extend the methodology used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
I fit a similar model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
Finally, I poststratify the posterior distributions of incidence to the CIS target population and subpopulations of interest.

@@

Having fit the prevalence model, I draw 10,000\todo{check number} prevalence trajectories in each stratum from the approximate posterior distribution of the model.
For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
Considering the relationship between incidence and prevalence to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies, equation implies that, for any stratum $\vec{P} = \matr{S} \vec{Z}$ where $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum, $\matr{S}$ is a matrix as defined below, and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
The matrix $\matr{S}$ is defined from the survival function to preserve the relationship between incidence and prevalence.
It is lower-triangular of the following form:
\begin{align}
    (\matr{S})_{i,j} = \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases}
\end{align}
Since the matrix is lower-triangular, the solution for $\vec{P}$ can be found efficiently using forward substitution\todo{cite forward substitution}.
Forward substitution is numerically stable unless the diagonal elements are close to 0.
Here, the diagonal is $S(1) = 1$ and hence the matrix is well-conditioned.
\todo{I explained this more clearly in my first-year report, look that up!}

@@

The transmission part of the model follows \textcite{birrellRealtime}.
A novel approach (\cref{SEIR:sec:observation}) links the transmission model to the data.
In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, it includes both the prevalence process and the observation process (as defined in \cref{E-inc-prev}).

@@

\subsection{Methods}
\todo[inline]{refer to where quantities are defined}
I use the same CIS data as in \cref{E-backcalc}, again considering each region independently.
The model will be stratified only by region and age group, so the data are aggregated to these levels.
Only individuals aged two or older are eligible to participate in the CIS; I assume that test results in all participants aged under 11 are still representative of the whole $[0, 11)$ age group.

@@

Studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

@@

The transmission model represents the infection process (introduced in \cref{E-inc-prev:sec:infection-process}).
A variety of different approaches can be taken to link a transmission model to observations.
In the context of CIS, this requires representing the prevalence and observation processes (introduced in \cref{E-inc-prev:sec:prevalence-process,E-inc-prev:sec:observation-process} respectively)
The method to do so is known as the observation model.
The formulation of this model is heavily dependent on the data available.
This means that there is a wide variety of models used.
This section  explains the model I propose for fitting to the CIS prevalence data.

@@

A much more efficient option is to approximate this convolution by adding compartments to the ODEs.
This approach also allows the infected, pre-positive compartment, which I denote $p_0$, to be represented.
$p_0$ increases at the same rate as individuals flow into $e_1$.
These individuals then transition into a set of compartment with their transition rates, $p_1, \dots, p_l$, chosen to match the duration of positivity estimates in \cref{E-imperf-test}.
These compartments do not have a biological interpretation, unlike the other compartments in the system, but are purely a mathematical tool so that the time spent in these compartments matches the duration distribution.
This is the same idea behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).

@@

The modelled proportion of individuals that are PCR positive in the are linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
In this parameterisation (detailed in \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.

@@

Priors for the parameters are given in \cref{SEIR:table:priors}.
The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors are used for them.
$\sigma_\epsilon$ and $\rho$ are given priors that penalise model complexity.
The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
The prior on $\rho$ favours a null model of a binomial likelihood.
The remaining priors had weakly informative priors.
